---
title: "QC report - Gestation and birth weight by gender"
author: "Hung Pham"
date: "04/21/2021"
output:
  pdf_document:
    number_sections: TRUE
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = T, warning = F, message = F)
```

# Synopsis
The "birth weight" data is a toy dataset that contains simulated information of 
135 babies. The variables recorded are birth weight (grams), length of gestation
(weeks), gender (Male/Female), and unique ids.

This report aims to accomplish two routine tasks: 1) Quality Control a dataset, 
and 2) produce preliminary statistical analyses. In this case, the statistical
analyses of interest is the association between birth weight, length of 
gestation, and gender.

Data is not uploaded to github. 

# Libraries
The following code chunk loads the relevant R libraries - collections of pre-
programmed R functions wrapped in discrete packages.
```{r}
library(tidyverse) # Data cleaning and wrangling
library(janitor) # Data cleaning
library(here) # For reproducibility and ease of collaboration
library(knitr) # Table formating 
library(kableExtra) # This library formats our table to look nice
library(rlang) # For writing functions
library(patchwork) # For visualization
library(GGally) # For pairwise plot
library(magrittr) # Piping
library(treemap) # Visualize categorical variables
library(xtable) # Latex tables
```

# QC
## Data input
- Import the data with `read_csv`. There are *135 obs* and *4 variables*.
- Add another column `unique_id` in case there are problems with the `new_id`
column.
- Convert to lower case and remove problematic 
symbols (e.g. `\`,`,`,`...`, etc.) from column names with `clean_names`.
- `my_dat` is our raw data. We print the first 10 observations from the data.

```{r}
loc<-here() # This variable tracks the root path of this file on a local machine

figure_n<-0 # Counter for figure number
count_fig<-function(reset = NULL){
        reset<- enexpr(reset)
        if(is.null(reset)){
        figure_n<<-figure_n+1} else {figure_n<<- 0}
} # This function automatically update our figure number everytime its ran

xtable2kable <- function(x) {
        out <- capture.output(print(x, table.placement = NULL))[-(1:2)]
        out <- paste(out, collapse = "\n")
        structure(out, format = "latex", class = "knitr_kable")
} # This function solves a bug with Latex

my_dat<-read_csv(paste0(loc,"/data/Play_Data_V1.csv")) %>% 
        clean_names() # Load the data into R 
        
my_dat$id<- 1:nrow(my_dat) # Add a unique id column.

head(my_dat, 10) %>% 
        xtable(caption = "Top 10 obs of raw data") %>% 
        xtable2kable() %>%
        kable_classic_2(latex_options = "HOLD_position", position = "center")
```

- Immediately, we see some issues:\
        - `gender` has numerical values.\
        - There are missing data in `birthweight` and `gender`.
- We first identify the problematic values in each variable. Then we will 
process missing data.

## Identify individual errors
### `birthweight`
```{r birth weight stat}
summary(my_dat$birthweight) %>% 
        broom::tidy() %>%  
        xtable(digits = 4, caption = "Summary stats of birth weights") %>% 
        xtable2kable() %>%
        kable_classic_2(latex_options = "HOLD_position",position = "center") 
```

`birthweight` is measured in grams. Value as low as 0.2983 or as 
high as 300033.5 are improbable. We print the `birthweight` values of < 5% 
percentile and > 95% percentile.

```{r outside of range}
bw_quantile<-quantile(my_dat$birthweight, c(0.05,0.95), na.rm = T)

my_dat %>% 
        filter(birthweight < bw_quantile[1] | birthweight > bw_quantile[2]) %>%
        arrange(birthweight) %>%   
        xtable(digits = 3, 
               caption = "Obs where birth weight are outside of the 5th-95th percent range") %>%
        xtable2kable() %>%
        kable_classic_2(latex_options = "HOLD_position",position = "center") 
```

Clearly, 0.2983, 0.3855, and 3e6 grams are not possible. We label and collect
these errors for fixing.

```{r}
problem_row<-tibble(id = c(23,34,69),
                    issue = c("bw too small", "bw too small", "bw too large"))
```

### `length_of_gestation`
```{r gastation length stats}
summary(my_dat$length_of_gestation) %>% 
        broom::tidy() %>%    
        xtable(digits = 3, caption = "Summary stats of gestation length") %>% 
        xtable2kable() %>% 
        kable_classic_2(latex_options = "HOLD_position",position = "center") 
```

For `length_of_gestation`, value as high as 380.1429 is improbable. We gather
values > 95% percentile.

```{r}
lg_quantile<-quantile(my_dat$length_of_gestation, 0.95, na.rm = T)
my_dat %>% 
        filter(length_of_gestation > lg_quantile[1]) %>%
        arrange(length_of_gestation) %>%    
        xtable(digits = 3, 
               caption = "Obs where gestation length are > 95th percentile") %>% 
        xtable2kable() %>%  
        kable_classic_2(latex_options = "HOLD_position",position = "center") 
```

We collect the id where length of gestation is 380.1429 and move on. 
```{r}
problem_row<- problem_row %>%
        bind_rows(tibble(id = 135,
                         issue = c("gestation too long")))
```

### `new_id`
```{r}
head(my_dat$new_id) # Top new_id values
class(my_dat$new_id) # Type of variable for new_id
```

The fact that the top values of `new_id` are all integers yet R imported 
`new_id` as a character variable raised concerns. We use REGEX to detect
which values are not integers.
```{r}
my_dat %>% 
        filter(!str_detect(new_id, "\\d")) %>%    
        xtable(digits = 3, caption = "Obs where id is not an integer") %>% 
        xtable2kable() %>%  
        kable_classic_2(latex_options = "HOLD_position",position = "center") 
```

IDs shouldn't be `Male` or `Female`. Also, individual IDs should be unique. 
We check if there are any repeats in `new_id`.
```{r}
my_dat %>% 
        filter(str_detect(new_id, "\\d")) %>% 
        pull(new_id) %>% 
        unique(.) %>% 
        length(.)
```

The math checks out. There are 133 unique numerical IDs after removing two 
wrong IDs. We confirm that each observation is unique and collect the 
problematic IDs rows.
```{r}
problem_row<- problem_row %>%
        bind_rows(tibble(id = c(26,45),
                         issue = c("wrong ID")))
```

### `gender`
```{r}
tabyl(my_dat,gender) %>%    
        xtable(digits = 3, caption = "Frequency table of gender variable") %>% 
        xtable2kable() %>%  
        kable_classic_2(latex_options = "HOLD_position",position = "center") 
```

Gender shouldn't be floats. We collect these rows.
```{r}
wrong_gender<- my_dat %>% 
        filter(str_detect(gender,"[[\\d]]")) %>% pull(id)

problem_row<- problem_row %>%
        bind_rows(tibble(id = wrong_gender,
                         issue = c("wrong gender")))
```

## Missing data
We investigate the pattern of missing values.
```{r, fig.align='center',fig.width=6,fig.height=5, cache= T}
missing_to_binary<-function(x){if_else(is.na(x),1,0)}

# `for_miss_viz` converts missing values to 1, non missing values to 0, and
# problematic values to 2. 
for_miss_viz<-my_dat %>% 
        mutate(across(.cols = -id,
                      .fns = missing_to_binary)) %>% 
        pivot_longer(cols = -id,
                     names_to = "column_name") %>% 
        left_join(problem_row, by = "id") %>% ### Join with the problem table
        mutate(value = case_when(
                ### Convert problematic values to 2
                grepl("bw too", .$issue) & column_name == "birthweight" ~ 2,
                grepl("gestation", .$issue) & column_name == "length_of_gestation" ~ 2,
                grepl("wrong ID", .$issue) & column_name == "new_id" ~ 2,
                grepl("gender", .$issue) & column_name == "gender" ~ 2,
                T ~ value),
               ### Convert data to factor for visualization
               x_coord = fct_inseq(as.character(rep(1:4, 135))),
               y_coord = fct_rev(fct_inseq(as.character(id))),
               my_color = factor(value))

count_fig() # Count figure
ggplot(for_miss_viz, aes(x_coord, y_coord)) +
        geom_tile(aes(fill = my_color),color = "white")+
        scale_fill_manual(values = c("grey","black","red"),
                          labels = c("not missing", "missing", "problematic value"))+
        scale_x_discrete(labels = setdiff(names(my_dat),"id"), position = "top")+
        theme_minimal()+
        labs(title = paste0("Figure ",figure_n,": Missing matrix of birth weight data"))+
        theme(axis.title = element_blank(),
              axis.text.y = element_blank(),
              panel.grid = element_blank(),
              legend.title = element_blank(),
              legend.position = "right")
```

From Figure 1, we conclude that **available data analysis** is appropriate. 
Meaning bias is not an issue if we proceed with the analysis only
using the available data. This is because the amount of missing data is not 
significant and the pattern can be safely concluded as missing at completely 
random (MCAR). Sensitivity analysis with imputed data using multiple imputation 
can be performed to confirm the robustness of the available data analysis but it wouldn't be needed in this scenario.

## Removing problemmatic values
We can setup a for-loop to remove the problematic values.
```{r}
tempt_dat<-for_miss_viz %>% filter(value == 2) %>% select(id,column_name)

my_dat_clean<-my_dat

remove_walk<-function(x,y){
        my_dat_clean[which(my_dat_clean$id == x), y]<<-NA
}

walk2(tempt_dat$id,tempt_dat$column_name,remove_walk)
```

We check the imputation.
```{r}
before_remove<-problem_row %>% left_join(my_dat,by = "id") %>% 
        rename(gestation_l = length_of_gestation)
after_remove<-problem_row %>% left_join(my_dat_clean,by = "id") %>%
        rename(gestation_l = length_of_gestation)

xtable(digits = 3, x =  before_remove,
       caption = "Obs where there are improbable values, before replacing the values with NA") %>%
        xtable2kable() %>%
        kable_classic_2(latex_options = "HOLD_position",position = "center")
xtable(digits = 3,x =  after_remove,
    caption = "After replacing the values with NA") %>%
        xtable2kable() %>%
        kable_classic_2(latex_options = "HOLD_position",position = "center")
```

This table shows the list of all our problematic values and the successful 
removal of those values. `my_dat_clean` is our cleaned data. 

# Statistical Analysis
## Birthweight and length of gestation
### Normality assumption
```{r}
## This function draws qq plots
draw_qq<-function(my_var,dat,strata = NULL){
        if(!is.null(strata)){
                p<-dat %>% ggplot(aes(sample = .data[[my_var]])) +
                        geom_qq()+
                        geom_qq_line()+
                        theme_bw()+
                        facet_wrap(~ .data[[strata]])+
                        labs(subtitle = paste0(my_var," qq-plot"))+
                        theme(plot.subtitle = element_text(hjust = 0, size = 13))
                return(p)
        }
        p<-dat %>% ggplot(aes(sample = .data[[my_var]])) +
                        geom_qq()+
                        geom_qq_line()+
                        theme_bw()+
                        labs(subtitle = paste0(my_var," qq-plot"))+
                        theme(plot.subtitle = element_text(hjust = 0, size = 13))
        p
}
p2<-draw_qq("birthweight",my_dat_clean)
p3<-draw_qq("length_of_gestation",my_dat_clean)
```

```{r, fig.align='center',fig.width=12,fig.height=6}
count_fig() # Update figure number
p2+p3+plot_annotation(title = paste0("Figure ",figure_n,
                                     ": qqplots of birthweight and gestation length"))
```

From the qq-plots, we see that length of gestation might not be normally 
distributed. We use the Shapiro-Wilk test to formally test the normal assumptions
of `birthweight` and `length_of_gestation`. 

```{r}
## This function performs the shapiro-wilk test and automatically generate 
## decisions when pvalue is >= 0.05

my_shapiro<-function(my_var,dat){
        for_map<-function(x){
                res<-shapiro.test(dat[[x]]) %>%
                        broom::tidy()
                # If p.value is larger than 0.05 then we reject the null hypothesis
                if (res$p.value >= 0.05) {
                        res %<>% mutate(H0 = paste0(x, " is normally distributed"),
                                        decision = "fail to reject H0",
                                        n = length(na.omit(dat[[x]]))) %>% 
                                select(statistic, p.value, n, everything())
                } else {
                        res %<>% mutate(H0 = paste0(x, " is normally distributed"),
                                        decision = "reject H0",
                                        n = length(na.omit(dat[[x]]))) %>% 
                                select(statistic, p.value, n, everything())
                }
        }
        res<-map(my_var,for_map) %>% reduce(rbind)
        res
}
my_shapiro(c("birthweight","length_of_gestation"), my_dat_clean) %>% 
        mutate(method = "Shapiro-Wilk") %>% 
        xtable(digits = 3, 
               caption = "Testing the normality assumption of birth weight and gestation length") %>% 
        xtable2kable() %>%  
        kable_classic_2(latex_options = "HOLD_position",position = "center") 
```

We reject the null hypothesis that length of gestation is normally distributed
and fail to reject the null hypothesis that birth weight is normally distributed.

### Correlation
```{r, cache=F, fig.align='center',fig.width=4,fig.height=4}
count_fig()
my_dat_clean %>% ggplot(aes(x = birthweight, y = length_of_gestation)) +
        geom_point()+
        geom_smooth(method = "loess")+
        labs(title = paste0("Figure ",figure_n,
                                     ": Scatterplot of birthweight by length of gestation"))+
        theme_bw()
```

From the scatter-plot, we can safely assume that there exists a linear correlation
between `birthweight` and `length_of_gestation` for the range of values in our 
data. Hence, we can use a Pearson's correlation to measure this correlation. 

```{r}
pearson_res<- cor.test(my_dat_clean$birthweight, my_dat_clean$length_of_gestation) %>% 
        broom::tidy() %>% rename(correlation = estimate) %>% 
        t() %>% as.data.frame()
        
pearson_res %>% tibble(term = rownames(.)) %>%
        rename("Value" = V1) %>% select(term,Value) %>% 
        kable(digits = 3, caption = "Pearson's correlation test") %>% 
        kable_classic_2(latex_options = "HOLD_position",position = "center") 
```

The Pearson's correlation between `birthweight` and `length_of_gestation` of 
0.622 *is significant* at a two sided alpha of 0.05.

### Differential effect by gender
We remake the qq-plots by gender using `draw_qq` programmed on line 257.
```{r}
p4<-draw_qq("birthweight",my_dat_clean,"gender")
p5<-draw_qq("length_of_gestation",my_dat_clean,"gender")
```

```{r, fig.align='center',fig.width=8,fig.height=6}
count_fig()
(p4/p5)+plot_annotation(title = paste0("Figure ",figure_n,
                                     ": qqplots of birthweight and gestation length by gender"))
```

#### Is the birthweight between males and females babies the same?
We use the t-test to test the following H0: *the mean `birthweight` between males and females are the same.*
```{r}
t_test_res<- t.test(my_dat_clean[which(my_dat_clean$gender == "Female"),
                                 "birthweight"],
       my_dat_clean[which(my_dat_clean$gender == "Male"),"birthweight"],
       var.equal = T) %>% 
        broom::tidy() %>% 
        rename("mean diff" = estimate,
               "mean female" = estimate1,
               "mean male" = estimate2) %>% t() %>% as.data.frame()

t_test_res %>% tibble(term = rownames(.)) %>% rename("Value" = V1) %>% 
        select(term,Value) %>% 
        xtable(digits = 3, 
               caption = "T-test mean difference between male and female birth weight, assuming
            equal variance") %>% 
        xtable2kable() %>% 
        kable_classic_2(latex_options = "HOLD_position",position = "center") 
```

#### Is the length of gestation between males and females babies the same?
We use the non-parametric Wilcoxson-rank sum test to test the following H0: 
*the location shift (or median) of `length_of_gestation` between male and female is zero.*
```{r}
wil_cox_test<-wilcox.test(my_dat_clean[which(my_dat_clean$gender == "Female"),"length_of_gestation", drop = T],
       my_dat_clean[which(my_dat_clean$gender == "Male"),"length_of_gestation", 
                    drop = T]) %>% 
        broom::tidy() %>% t() %>% as.data.frame() 

wil_cox_test %>% tibble(term = rownames(.)) %>% rename("Value" = V1) %>% 
        select(term,Value) %>% 
        xtable(digits = 3,
               caption = "Wilcox test of median shift between male and female length of gestation") %>% 
        xtable2kable() %>% 
        kable_classic_2(latex_options = "HOLD_position",position = "center") 
```

- We fail to reject the null hypothesis that the mean `birthweight` is the same
between male and female with a two sided alpha = 0.05.

- We fail to reject the null hypothesis that the median of `length_of_gestation` is
the same between males and females with a two sided alpha = 0.05.

# Recode
We create another variable "term" where `length_of_gestation` < 37 weeks is 
categorized as "pre-term" and "to term" otherwise. Then we create a treemap to
visualize the proportion of to-term and pre-term by gender. 
```{r}
my_dat_clean$term<-ifelse(my_dat_clean$length_of_gestation < 37,
                           "pre-term",
                           "to term")

tree_dat<-my_dat_clean %>% 
        group_by(term, gender) %>% 
        summarize(n = n()) %>% 
        rbind(my_dat_clean %>% 
                      group_by(term) %>%
                      summarize(n = n()) %>% 
                      mutate(gender = "both")
        )
tree_dat %>% 
        kable(digits = 3, caption = "Frequency table of `term` alone and by gender") %>% 
        kable_classic_2(latex_options = "HOLD_position",position = "center") 
```

```{r, fig.align='center',fig.width=7,fig.height=5}
treemap(tree_dat[1:6,],
        index = c("term", "gender"),
        vSize = "n",
        title = "Treemap of gestation term by gender")
```

# Session Info
Session information for reproducibility.
```{r}
sessionInfo()
```

